PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-06-19 manuscript_00.Rmd
updates

diff_tab %>% 
  group_by(Base, Organism) %>% 
  summarise(Count = n()) %>% 
  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
  as_tibble() %>% 
  flextable()

# number of celltype (predict) - organism - study counts (with >=50 cells)
meta_filter %>%  
  filter(Organ == 'Eye') %>% 
  group_by(organism, CellType_predict, study_accession) %>% 
  count() %>% 
  filter(n >= 50) %>% 
  ungroup() %>% 
  group_by(organism, CellType_predict) %>% 
  count() %>% 
  pivot_wider(values_from = n, names_from = organism) %>% 
  filter(!is.na(CellType_predict)) %>% 
  replace(is.na(.), values = 0) %>% 
  arrange(CellType_predict) %>% 
  rename(`CellType (Predict)` = CellType_predict) %>% 
  mutate(`Well Supported` = case_when(`Homo sapiens` > 2 & `Mus musculus` > 2 ~ 'Yes',
                                      TRUE ~ 'No')) %>% 
  flextable(set_caption())
```

@@

We filter down to only assess cell types with more than two independent datasets in both human and mouse. This leaves ..... These well supported cell types can be used to more strenuously identify genes that are consistently differentially expressed in a cell type across multiple species. To define these well supported marker genes we only ass
```{r}
ws_oct <- meta_filter %>%  
  filter(Organ == 'Eye') %>% 
  group_by(organism, CellType_predict, study_accession) %>% 
  count() %>% 
  filter(n >= 50) %>% 
  ungroup() %>% 
  group_by(organism, CellType_predict) %>% 
  count() %>% filter(n >= 3) %>% 
  ungroup() %>% 
  group_by(CellType_predict) %>% 
  summarise(Count = n()) %>% 
  filter(Count > 1, !is.na(CellType_predict)) %>% 
  pull(CellType_predict)
```

@@

```{r}
consist_diff <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Base %in% ws_oct) %>%  
  filter(Group == 'CellType (Predict)', 
         Against == 'All') %>% 
  group_by(Base, Gene) %>% 
  summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-4), `mean padj` = mean(padj)) %>% 
  filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%  
  collect() %>% 
  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
  left_join(gene_tab, by = 'GENEID') %>% 
  select(-Gene) %>% 
  relocate(SYMBOL, GENEID) %>% 
  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
  arrange(-`mean log2FC`)

@@

### heatmap human
```{r, fig.width=6, fig.height=17}
# diff_ratios <- diff_tab_big %>% 
#     filter(Against != 'All') %>% 
#     group_by(Symbol, ID, Organism, Base) %>% 
#     summarise(Count = n()) %>% 
#     filter(Count > 1) %>% 
#     left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
#     mutate(Ratio = `Count.x` / `Count.y`) %>% 
#     select(Symbol:Base, Ratio)

gene_sets <- diff_tab %>% 
  #left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
  #filter(Ratio > 0.2) %>% 
  group_by(Base, Organism) %>% 
  slice_max(n=5, order_by = log2FoldChange)

org <- 'Homo sapiens'

g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()

# only uses CT that exist in the organism <-> CellType 
ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()

x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Gene %in% g) %>% 
  collect() %>% 
  filter(Base %in% ct,
         Group == 'CellType (Predict)', 
         Organism == org, Against == 'All') %>% 
  select(Gene, log2FoldChange, Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene
x <- x[,-1]
Heatmap(x, column_labels = celltype)
```

### heatmap mouse
```{r, fig.width=5, fig.height=20}

org <- 'Mus musculus'

g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()

# only uses CT that exist in the organism <-> CellType 
ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()

x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Gene %in% g) %>% 
  collect() %>% 
  filter(Base %in% ct,
         Group == 'CellType (Predict)', 
         Organism == org, Against == 'All') %>% 
  select(Gene, log2FoldChange, Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene
x <- x[,-1]
Heatmap(x, column_labels = celltype)
```

### heatmap macaque
```{r, fig.width=5, fig.height=20}

org <- 'Macaca fascicularis'

g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()

# only uses CT that exist in the organism <-> CellType 
ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()

x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Gene %in% g) %>% 
  collect() %>% 
  filter(Group == 'CellType (Predict)', 
         Organism == org, Against == 'All', Base %in% ct) %>% 
  select(Gene, log2FoldChange, Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene
x <- x[,-1]
Heatmap(x, column_labels = celltype)
```

>>>>

diff_tab_sup <- diff_tab %>% 
  group_by(Base, Organism) %>% 
  summarise(Count = n()) %>% 
  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
  as_tibble()
write_csv(diff_tab_sup, file = 'supplemental_files/differential_expression_table.csv')
diff_tab_sup %>% 
  flextable()

# number of celltype (predict) - organism - study counts (with >=50 cells)
meta_filter %>%  
  filter(Organ == 'Eye') %>% 
  group_by(organism, CellType_predict, study_accession) %>% 
  count() %>% 
  filter(n >= 50) %>% 
  ungroup() %>% 
  group_by(organism, CellType_predict) %>% 
  count() %>% 
  pivot_wider(values_from = n, names_from = organism) %>% 
  filter(!is.na(CellType_predict)) %>% 
  replace(is.na(.), values = 0) %>% 
  arrange(CellType_predict) %>% 
  rename(`CellType (Predict)` = CellType_predict) %>% 
  mutate(`Well Supported` = case_when(`Homo sapiens` > 2 & `Mus musculus` > 2 ~ 'Yes',
                                      TRUE ~ 'No')) %>% 
  flextable()
```

@@

We filter down to only assess cell types with more than two independent datasets in both human and mouse. This leaves ..... These well supported cell types can be used to more strenuously identify genes that are consistently differentially expressed in a cell type across multiple species. To define these well supported marker genes we only ass

@@

```{r}
ws_oct <- meta_filter %>%  
  filter(Organ == 'Eye') %>% 
  group_by(organism, CellType_predict, study_accession) %>% 
  count() %>% 
  filter(n >= 50) %>% 
  ungroup() %>% 
  group_by(organism, CellType_predict) %>% 
  count() %>% filter(n >= 3) %>% 
  ungroup() %>% 
  group_by(CellType_predict) %>% 
  summarise(Count = n()) %>% 
  filter(Count > 1, !is.na(CellType_predict)) %>% 
  pull(CellType_predict)

consist_diff <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Base %in% ws_oct) %>%  
  filter(Group == 'CellType (Predict)', 
         Against == 'All') %>% 
  group_by(Base, Gene) %>% 
  summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-4), `mean padj` = mean(padj)) %>% 
  filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%  
  collect() %>% 
  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
  left_join(gene_tab, by = 'GENEID') %>% 
  relocate(SYMBOL, GENEID) %>% 
  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
  arrange(-`mean log2FC`)

@@
